\section{Examples}

\newcounter{example}[section]
\newenvironment{example}{\refstepcounter{example}\paragraph{Example \theexample:}}{}

\begin{example}
You are a mage who is near the end of an arduous journey through a dungeon, looking for a golden teapot at the end of it.
You open the door to the final room and enter into a long, narrow room largely taken up by a gaping abyss.
On the other side of the abyss you see the teapot!
You fish through your pack for your Scroll of Magical Bridge.
Your level in the Construction Magic ability is $6$,
but the GM determines
$D = 7$
by considering that a standard being with $7$ levels worth of Construction Magic XP would succeed half the time at casting this spell over an abyss of that size.
Additionally, you recently fell ill after eating the wrong kind of dungeon mushroom,
and this gave a property ``Shroom Belly'' that makes it harder to focus on magic,
granting $A = -1$ for magical actions.
You roll $4$dF... and you get $+3$!
$L = 6$, $A = -1$, $R = +3$, $D = 7$, and $8 \geq 7$.
You successfully create the bridge, walk across it, and grab the golden teapot!
The GM awards you $10$ XP for completion of this narrative chunk.
You decide to allocate $7$ XP towards Construction Magic, and $3$ XP towards Mycology (another ability you used in this dungeon, albeit not successfully).
Construction Magic already had $22$ XP allocated to it (out of the $64$ XP required to raise it to level $7$), and now has $29$.
Mycology was at level $0$, and has now been raised to level $2$.
\end{example}


\begin{example}
(placeholder for setup example)
\end{example}


\begin{example}
You are the GM and the players have just fought their way into a dungeon room with a closed door at the opposite end.
The door is cracked open from the other side, just barely revealing a stack of nasty gremlins.
Adam announces that his archer character attempts to fire an arrow into the narrow crack.
As the GM, you need to set the difficulty $D$ for this action.
You decide that in order to have about a $50\%$ chance of success, one would need about a year of dedicated archery training.
Earlier in the campaign a different character did a month of dedicated training and received $2$ XP for her archery ability.
A year of dedicated archery training then corresponds to $24$ XP, which corresponds to a level of about $4$.
So you let Adam know that $D=4$ for this action.
``That’s pretty harsh man,'' Adam complains.
``It’s not that hard for an archery task.''
Being the wise GM that you are, you understand that $D$ comes down to the amount of needed practice and not a comparison with similar tasks.
Once you explain how you determined $D$, Adam has a further complaint:
``My archery ability level is $2$, which corresponds to $3$ XP.
By your convention that’s $1.5$ months of archery practice… but my character has been an archer for like $3$ months!
What gives?!
Why do I get less XP?!'' After calming Adam down, you explain that his character’s practice wasn’t dedicated archery training.
His $3$ months of undedicated practice is being treated as though it were $1.5$ months of dedicated training.
Finally, Adam needs to roll to resolve the action.
With $L=2$ and $D=4$, it appears that Adam has to roll at least $+2$ to succeed.
Rolling $4$dF, this gives him about a $1/5$ chance of success.
Adam rolls his $4$dF and gets $R=-1$, failing the shot.
Since $-1$ is not such a surprising roll, you do not interpret Adam’s failure in a dramatically unlucky way.
His relative performance, on the other hand, is quite low: $L+A+R-D = 2+0-1-4 = -3$.
And so you narrate that the arrow appears to have had little hope of hitting its target,
and the gremlins all snicker at Adam’s character.
\end{example}


\begin{example}
Three characters, Gromanno (PC), Nyrv (NPC), and Polthova (PC) just came across a set of exquisite marbles in a cave, and they all want it.
They decide to race to the entrance of the cave, and the marbles will go to the winner.
The GM says that this will be treated as an opposed action,
where $L+A+R$ is compared among all participants, and the ability being used is \emph{Fitness}.
In this campaign,
a Fitness of $2$ is interpreted as poor,
a Fitness of $4$ is interpreted as decent,
and a Fitness of $7$ is interpreted as excellent.
Here is the outcome for $L$, $A$, and $R$:
\begin{itemize}
\item
$L$: Gromanno has a Fitness of $2$, Nyrv has a Fitness of $5$, and Polthova is a professional athlete with a Fitness of $8$.
\item
$A$: Gromanno is wearing his ``Old running shoes [$+1$ to $A$ for running actions]'' and the GM gives him an $A$ of $+1$ for this.
Nyrv has the property ``Poor low-light vision'' and can’t navigate well in the cave, and the GM gives her an $A$ of $-1$ for this.
Polthova does not have any distinct character-based factors that would affect his $A$ for this action, and the GM gives him an $A$ of 0.
\item
$R$: They all roll $3$dF.
Gromanno’s player rolls a $0$ for him, the GM rolls a $+2$ for Nyrv, and Polthova’s player rolls a $-3$ for him.
\end{itemize}
Comparing $L+A+R$ for all participants:
\begin{itemize}
\item Gromanno: $2+1+0 = 3$
\item Nyrv: $5-1+2 = 6$
\item Polthova: $8+0-3 = 5$
\end{itemize}
The GM narrates the events that led to this outcome.
Polthova, very fit, started out ahead, with the other two trailing far behind him.
But due to his terrible luck, around two thirds of the way through, he slipped on some cave slime and injured his ankle.
He limped the rest of the way to the entrance.
Nyrv exerted herself very well and was able to push through her difficulty navigating, and eventually surpassed the limping Polthova.
Gromanno, as expected, trailed behind the other two, despite his nice running shoes.
Nyrv emerges from the cave first, victorious, delighted in her new set of exquisite marbles!
\end{example}


\begin{example}
In a campaign where the GM has decided that humans are the standard being for the Strength ability,
and that a Strength level of 3 is to be interpreted as ``decent,'' an odd group of pals is sitting around a table arm wrestling.
One is a hobbit, one is a cyborg human with a mechanical left arm, and one is an ant bodybuilder.
\begin{itemize}
\item The hobbit, small by nature, has an unmodified Strength of $3$ (for the purposes of determining XP cost for leveling up) and a property
``hobbit strength'' that modifies his Strength by $-1$ (for the purposes of using Strength in an action); his i$L$ is therefore represented as $3-1=2$.
\item The cyborg human has an unmodified Strength of $3$, and a property ``cyborg left arm'' that modifies his Strength by $+2$ when using the arm in an action;
his $L$ for Strength for actions using his left arm can be represented as $3+2=5$, and for actions not using his left arm would be represented as $3+0=3$.
\item The ant, an avid athlete but several orders of magnitude smaller than its friends, has a property ``bug strength'' that modifies its Strength by $-50$.
Its $L$ is represented as $7-50=-43$.
\end{itemize}
The GM and players agree that $A$ for all characters is $0$ here, with no current situational factors that will aid or hold back any characters.

The GM decides to resolve the arm wrestling match as a series of opposed actions.
He believes that this is an action that incorporates less luck than most actions (where the better prepared character is more certain to succeed than usual),
and decides that $1$dF is to be used for $R$.

The group agrees to using their left arms (left foreleg in the case of the ant), as is customary in this world.
The hobbit and the cyborg face off first:
the hobbit rolls a $+1$ for an $L+A+R$ of $2+0+1=3$, and the cyborg rolls a $-1$ for an $L+A+R$ of $5+0-1=4$.
After a long back-and-forth struggle, the cyborg wins.

Next, the hobbit and the ant face off.
The hobbit’s arm is tired after the last match, and $A$ for him will be $-1$ for this next match.
The hobbit rolls a $-1$ for an $L+A+R$ of $2-1-1=0$, and the ant rolls a $+1$ for an $L+A+R$ of $-43+0+1=-42$.
Despite the ant’s noble preparation and efforts, it is flung across the room immediately.

The pals schedule a rematch, and each of them resolves to train in preparation.
\begin{itemize}
\item
The hobbit lifts weights for a few minutes every day.
After a few weeks of training,
the GM awards the hobbit $9$ XP to be allocated to Strength, raising his unmodified Strength to $4$ ($+\frac{1}{16}$ towards the next level).
His $L$ would now be represented as $4-1=3$.
\item
The cyborg human repeatedly exerts his left arm in a misguided attempt to apply biological strength training concepts to a mechanical arm.
No XP is gained.
\item
The ant is determined to level up its Strength.
The ant’s current unmodified Strength is $7$, so the XP needed for a level-up is $2^7$, and the ant trains accordingly.
The ant’s effective Strength for the next match has increased to $L=8-50=-42$.
\end{itemize}
\end{example}
