\section{Campaign Setup Checklist}
The following checklist only focuses on the \peupfudge{} aspects of a campaign that should be ready before the first session.
It omits elements of worldbuilding such as maps and lore,
and it omits setup that is done on a per-session basis such as rooms and NPC character sheets.

\def\itemSquare{\item[$\square{}$]}

\begin{itemize}
\itemSquare
Which characteristics should show up on character sheets?
\itemSquare
What kinds of beings could PCs be?
\itemSquare
What is the range of possible origin stories?
\itemSquare
Prepare the set of abilities.
We recommend first thinking about the sorts of actions characters might attempt, and then working backwards from there to find suitable abilities.
Estimate a level-proficiency relationship for each ability.
For example, you could record the levels for each ability that are considered to be poor, decent, excellent, etc.
\itemSquare
Prepare the set of statuses.
For each one, decide what states it can take, what causes the state to change, and what effect each state has.
\itemSquare
Are there any important properties that will serve to modify ability levels?
Note that the types of characters that do not end up with such properties are standard beings for the purpose of ability level interpretation.
\itemSquare
What is the starting XP and how might it vary between characters?
\itemSquare
Determine the procedure for populating the initial inventory and setting the initial states of statuses.
\itemSquare
Make a character sheet template.
\end{itemize}
